---
title: "Gov 50 Final Project"
author: Gregor Kossmann
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
## Introduction
When the European Recovery Program -- later dubbed the "Marshall Plan" -- was announced on the steps of tercentenary theater at Harvard on June 5, 1947, many understood its unprecedentedly vast scope but few could truly grasp the lasting and transformative effect it would have on reviving a war-torn European continent. The Marshall Plan was (and continues to be) America's most expansive foreign aid program, costing the US approximately 5% of its GDP in 1948 -- an unthinkably large expense today where foreign aid peaks at 1% of GDP. While this significant amount of money undoubtedly impacted Europe's economies and industries, I want to explore to what extent that holds true. In this paper I will be assessing the impact of the Marshall Plan on Europe's economy to ultimately determine how effective foreign aid is as a foreign policy tool. This is an especially relevant conversation in today's climate where the merits of aid to war-torn Ukraine are increasingly being debated. Thoroughly understanding historical instances of foreign aid that were successful is going to be critical in informing our perspectives today on what the most effective course of action would be to rebuild eastern Europe. Using publicly available data from Statista on the level of US loans to various European countries during the Marshall Plan period, through a regression model I will test how related the size of the loan to a country (the independent variable) is to various economic indicators (including industrial and agricultural output, which will be my dependent variables). Standard economic theory would suggest that given the massive influx of aid and capital that entered an already barren and deprived landscape in Europe, the Marshall Plan provided much needed stimulus to reinvigorate the economy. Therefore my hypothesis would state that those countries that received the most aid would also experience the biggest economic improvements. If, however, the patterns in the data show that the early years following the implementation of the Marshall Plan saw relatively meager growth, the null hypothesis would state that the Marshall Plan had no significant effect and that economic improvements in future decades were more likely the result of other policies (such as the Mutual Security Pact of 1951 that eventually replaced the Marshall Plan). 


## Figure 1: German Income in the Mid-20th Century
```{r, echo = FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(infer)
library(knitr)
library(kableExtra)
ZA8155_16 <- read_excel("~/Desktop/Gov50Data/ZA8155_16.xls")

real_inc <- as.numeric(as.character(ZA8155_16[["...3"]]))

german_inc <- ZA8155_16 |>
  summarize(`Reales Pro-Kopf-Einkommen in Europa und den USA (1870-1992)`, real_inc)|>
  slice(-(1:75)) |>
  slice_head(n = -35) |>
  ggplot(aes(x = `Reales Pro-Kopf-Einkommen in Europa und den USA (1870-1992)`, y = real_inc, fill = real_inc)) +
  geom_bar(stat = "identity", fill = "dodgerblue") +
  labs(x = "Years", y = "Income Per Capita in Germany (in mil. USD)") +
  scale_y_continuous(limits = c(0, 9000)) +
  geom_vline(xintercept = "1948", color = "indianred", linetype = "dashed") +
  geom_vline(xintercept = "1952", color = "indianred", linetype = "dashed") 
   
german_inc 
```
## Table 1: Change in US Loans (w/ Calculations)
```{r, echo = FALSE}
library(readxl)
US_Loans <- read_excel("~/Desktop/Gov50Data/Gov50USLoans1946-1961.xlsx", 
    col_types = c("text", "numeric", "numeric", 
        "skip"))

percentage_change <- ((US_Loans$"Marshall Plan period: 1949-52" - US_Loans$"Post-war relief period: 1946-1948") / US_Loans$"Post-war relief period: 1946-1948") * 100 

real_change <- (US_Loans$"Marshall Plan period: 1949-52" - US_Loans$"Post-war relief period: 1946-1948")

US_Loans <- US_Loans|>
  mutate("Percentage Change" = percentage_change) |>
  mutate("Real Change (in millions of US dollars)" = real_change)


US_Loans_sorted <- US_Loans[order(US_Loans$"Real Change (in millions of US dollars)", decreasing = TRUE), ]

knitr::kable(US_Loans_sorted, digits = 2)

```

## Figure 2: Change in US Loans
```{r, fig.width = 12, echo = FALSE}

real_change_graph <- US_Loans_sorted |>
  ggplot(aes(x = reorder(`U.S. overseas loans and grants to countries in Western Europe during various periods following the Second World War between 1946 and 1961 (in millions of U.S. dollars)`,
                         -`Real Change (in millions of US dollars)`),
             y = `Real Change (in millions of US dollars)`,
             fill = `Real Change (in millions of US dollars)`)) +
  geom_bar(stat = "identity")+
  coord_flip()+
  labs(x ="Countries Receiving Loans from US", y = "Change in Loan Amount in Marshall Period (1949-52), compared to 1946-48 levels")

real_change_graph
```

## Figure 3: Change in Agricultural Output
```{r, echo = FALSE}

Agdata <- read_excel("~/Desktop/Gov50Data/Gov50AgOutput1947-1951.xlsx", 
col_types = c("text", "numeric", "numeric", 
"numeric"))


Ag_perc_change <- ((Agdata$`1950-1951` - Agdata$`1946-1947`) / Agdata$`1948-1949`)*100  

Agdata <- Agdata |>
  mutate(perc_change = Ag_perc_change)

Agdata_sorted <- Agdata[order(Agdata$"perc_change", decreasing = TRUE), ]

Agdata_graph <- Agdata_sorted|>
  ggplot(aes(x = reorder(`Agricultural output in Western Europe 1947-1951`, - `perc_change`), y = `perc_change`, fill = `perc_change`)) + 
  geom_bar(stat = "identity")+
  coord_flip()+
  labs(y = "Percentage Change in Agricultural Output from 1947-1951", x = "")+
  scale_fill_continuous(name = "Percentage Change")

Agdata_graph

```

## Figure 4: Change in Industrial Output
```{r, echo = FALSE}
Industrydata <- read_excel("~/Desktop/Gov50Data/Gov50IndustrialOutput1947-1951.xlsx", 
col_types = c("text", "numeric", "numeric", 
"numeric"))

ind_perc_change <- ((Industrydata$`1951` - Industrydata$`1947`) / Industrydata$`1947`)*100  

Industrydata <- Industrydata |>
  mutate(perc_change = ind_perc_change)

Industrydata_sorted <- Industrydata[order(Industrydata$"perc_change", decreasing = TRUE), ]


Industry_graph <- Industrydata_sorted|>
  ggplot(aes(x = reorder(`Industrial output in Western Europe 1947-1951`, - `perc_change`), y = `perc_change`, fill = `perc_change`)) + 
  geom_bar(stat = "identity")+
  coord_flip()+
  labs(y = "Percentage Change in Industrial Output from 1947-1951", x = "")+
  scale_fill_continuous(name = "Percentage Change")

Industry_graph
```

## Regression Calculations
```{r, echo = FALSE}

common_countries <- intersect(intersect(US_Loans$`U.S. overseas loans and grants to countries in Western Europe during various periods following the Second World War between 1946 and 1961 (in millions of U.S. dollars)`, Industrydata$`Industrial output in Western Europe 1947-1951`), Agdata$`Agricultural output in Western Europe 1947-1951`)

US_Loans_subset <- US_Loans[US_Loans$`U.S. overseas loans and grants to countries in Western Europe during various periods following the Second World War between 1946 and 1961 (in millions of U.S. dollars)` %in% common_countries, ]
Industrydata_subset <- Industrydata[Industrydata$`Industrial output in Western Europe 1947-1951` %in% common_countries, ]
Agdata_subset <- Agdata[Agdata$`Agricultural output in Western Europe 1947-1951` %in% common_countries, ]

merged_data <- merge(US_Loans_subset, Industrydata_subset, by.x = "U.S. overseas loans and grants to countries in Western Europe during various periods following the Second World War between 1946 and 1961 (in millions of U.S. dollars)", by.y = "Industrial output in Western Europe 1947-1951")
merged_data <- merge(merged_data, Agdata_subset, by.x = "U.S. overseas loans and grants to countries in Western Europe during various periods following the Second World War between 1946 and 1961 (in millions of U.S. dollars)", by.y = "Agricultural output in Western Europe 1947-1951")

model_industrial <- lm(perc_change.x ~ `Real Change (in millions of US dollars)`, data = merged_data)
model_agricultural <- lm(perc_change.y ~ `Real Change (in millions of US dollars)`, data = merged_data)


summary_industrial <- summary(model_industrial)
summary_agricultural <- summary(model_agricultural)


table_industrial <- data.frame(
  Coefficients = rownames(summary_industrial$coefficients),
  Estimate = summary_industrial$coefficients[, "Estimate"],
  `Std. Error` = summary_industrial$coefficients[, "Std. Error"],
  `t value` = summary_industrial$coefficients[, "t value"],
  `Pr(>|t|)` = summary_industrial$coefficients[, "Pr(>|t|)"]
)

table_agricultural <- data.frame(
  Coefficients = rownames(summary_agricultural$coefficients),
  Estimate = summary_agricultural$coefficients[, "Estimate"],
  `Std. Error` = summary_agricultural$coefficients[, "Std. Error"],
  `t value` = summary_agricultural$coefficients[, "t value"],
  `Pr(>|t|)` = summary_agricultural$coefficients[, "Pr(>|t|)"]
)


kable(table_industrial, "html", align = "c") |>
  kable_styling(full_width = FALSE)

kable(table_agricultural, "html", align = "c") |>
  kable_styling(full_width = FALSE)

```



## Initial Observations
Comparing the data available, there are some rather intuitive conclusions that can be drawn. When looking at the income levels per capita in Germany (Figure 1), their steady rise during the Marshall Plan period (the values between the two red lines) is commensurate with the relatively large amounts of loan payments Germany is receiving from the US during that same period. Without doing complicated statistical analyses or complex calculations, we can intuitively see how significant aid to Germany is correlated with income per capita rising. Moreover, to support this observation it is also worth noting that other countries' aid plays a role in German income too (Table 1). Since other countries trade with each other, the amount of aid they receive should also factor in when determining conclusions. When performing brief calculations:

```{r}
avg_perc_finite <- mean(percentage_change[is.finite(percentage_change)])
avg_perc_finite

avg_real_finite <- mean(real_change[is.finite(real_change)])
avg_real_finite
```

we see that on average, European countries got approximately 6517.16% more in aid and loans from the US during the Marshall Plan (as compared to immediately following WW2). This figure excludes certain countries that received no aid at all before the Marshall Plan and therefore had a mathematically infinite percentage increase in aid. To put it more tangibly, on average, European countries during the Marshall Plan received approximately $738.92 million US dollars (this figure is somewhat more representative than the average percentage change as it includes more countries). These staggering figures, when put into perspective with the rising income levels in Germany, support the conclusion that the Marshall Plan had a demonstrable impact on the German economy. 

Despite these brief calculations, it would be somewhat disingenuous and statistically un-rigorous to end the conversation there, because crucially, the counter-factual is difficult to disprove. With the null hypothesis suggesting that the Marshall Plan yielded no significant improvements on the European economy, the data presented hitherto does not exactly disprove that. Although we might intuitively think that without such a huge influx of aid, income would not rise, we have not yet statistically proven that, and therefore need to conduct more tests. In the following section I will analyze the data from changes in the levels of industrial and agricultural output (Figure 3, Figure 4) for various European countries during the Marshall Plan years, and see whether the countries that received the most aid (as seen in Figure 2) were also the ones that saw biggest improvements in their agricultural and industrial output. These further tests (Regression Calculations) will strengthen the statistical validity of any conclusions from this paper, and genuinely explore whether the Marshall Plan significantly improved the European economy. 


## Analysis of Results 
Looking at the changes in agricultural output as well as industrial output in various European countries (Figure 3, Figure 4) is very helpful in terms of both providing a hollistic picture of the economy and directly relating to the hypothesis. If the hypothesis holds true then countries with a higher percentage of aid from Table 1 (such as France), should experience the biggest changes in agricultural and industrial output. However when taking a look at the graphs we see that is not actually happening. Countries like Belgium which got the fourth highest increase in aid from the US, ranked fifth in terms of agricultural output changes and eighth in terms of industrial output changes. Instead, we see countries like Austria -- which got the 11th highest increase in aid from the US -- increase the most in their agricultural output (38% increase) and the most in industrial output (196% increase), when compared to pre-Marshall Plan levels. What these observations suggest is that rather than aid being blanketly helpful, it was most helpful in recuperating countries that were most hit (i.e. Austria was heavily damaged during Allied counteroffensives). The regression model that was ran is very informative in determining how statistically significant many of these findings are, even if the output may be somewhat confusing. 

The intercept estimate ("(Intercept)") -- in the case of industrial output it was 48.55, and in the case of agricultural output it was 17.65 -- determines the estimated value of the dependent output (percentage change in output) when the independent variable (in this case the size of the loan and amount of aid) is zero. The slope ("Real Change (in millions of US dollars)") represents the estimated change in the dependent output (percentage change in output) for a one-unit change in the independent variable (one-unit increase in the size of the loan). For the industrial output the slope is 0.0006, and for agricultural output the slope is 0.0033, which tells us that on average, a one-unit increase in the size of US loans, leads to a 0.0006 increase in industrial output and 0.0033 increase in the agricultural output. The standard error and t-value are other pieces of information regarding the statistical validity of this regression, however they are not most relevant to the question at hand, which will primarily rely on the R-squared and p-value. The R-squared value -- which measures how much of the variance in the DV is explained by the IV -- for the agricultural output is 0.1118, while the R-squared value for the industrial output is 0.0003. Simply put, with a R-squared of 0.1118, it means that approximately 11% of the variance in the industrial output can be explained by the model (which is much higher than the 0.03% for the industrial data). On the other hand, the agricultural p-value is 0.264, while the industrial p-value is 0.952 The p-value is used to determine how the data stacks up against the null hypothesis, with a p-value of <0.05 suggesting it is safe to reject the null hypothesis. In this instance, despite the efforts to clarify the counterfactual, both industrial and agricultural data have extremely high p-values suggesting they are not meaningful predictors. 

All this jargon aside, what the regression proved to us was that there is not necessarily a definitive relationship between the amount of aid a country receives and increases in their economic output (whether it be industrial or agricultural). Therefore, this confirms the previous conclusion that aid does not have to be overwhelming to be effective. It just has to be tactically allocated where it makes most sense and where it might have the most impact. Clearly pouring hundreds of million of dollars into Portugal yields little results, but when even small amounts were given to countries like Austria (who needed it most), the results were most tangible. Ultimately this suggests that US aid is most effective when it is strategic and targeted, consequently proving to also be much more of an affordable endeavor. When linking back to the original hypothesis and research question, we see that the question is much more holistic than a simple binary of whether the Marshall Plan aid significantly helped Europe or not. Of course it did. The question now is where it helped most and how that can be more targeted in the future. 

## Limitations on Experiment
While I had the privilege of working with a relatively large data set, much of it was quite disjointed given the historical nature of it, forcing me to use data from different sources that had different methods of measuring and writing down the data. Even though I tried to standardize my data as much as possible, it posed quite an obstacle in this experiment. Furthermore, with more time, I would have liked to perform more advanced calculations on larger data sets that would hopefully yield more statistically rigorous conclusions. The present regression model I crafted is helpful in ruling out certain possibilities, however it is not satisfyingly definitive. I would also add that the data itself, despite it being visualized in graphs, itself is not very intuitive, and understanding it requires some level of historical knowledge (i.e. to discern that Austria's anomalous improvements were due to its geopolitical standing by the end of WW2), making it not the most accessible. Another limitation of this experiment was from the outset having to wrestle with using "real change" in US loans rather than "percentage change" in US loans, as my independent variable. While it would likely have been fairer to judge according to percentage change (since 500 million dollars goes much further in Norway than it does in France), I had to resort to "real change" because there was more data available and it did not discount the infinite values (since some countries went from no aid to some aid, the percentage change was thus definitionally infinite or undefined). 

Despite these shortcomings, they offer important reflections on how this experiment can be improved in the future with more time and resources (primarily through more sophisticated statistical calculations). Nevertheless, the regressions demonstrate that this data is highly statistically insignificant, pointing to the difficulty of assessing a genuine regression between such large (and historic) data sets. The key insights and takeaways remain the same though, US aid appears to be most effective when it is targeted to countries that would benefit most from it, rather than merely pouring overwhelmingly large amounts of money into certain economies. A fruitful further experiment would explore how to identify those countries that would benefit the most from this targeted aid, in order to inform lawmakers how to make foreign aid a more effective tool.   


# R Code {data-nav="collapsed"}
```{r, eval=FALSE}
knitr::purl("index.Rmd", "output_script.R")
cat(readLines("output_script.R"), sep = "\n")
```